Finally Tagless, Partially Evaluated

### 2008-09-12 tagless/tagless-final.tex
Add code using de Bruijn indices

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
In contrast to \S\ref{intro}, in the rest of the paper 
we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensures that our object programs are closed.
We find HOAS to be more convenient, but we have also implemented our
approach using de Bruijn indices (in the accompanying source file
|incope-dB.ml|).

@@

@@

Our encoding of the type function |repr_pe| in \S\ref{S:PE-solution}
emulates type-indexed types and is related to intensional type analysis
\citep{Morrisett-intensional,Generic-Haskell}. However, our object
language and running examples in HOAS include |fix|,
which intensional type analysis cannot handle
\citep{xi-guarded}.  Our final approach
seems related to \citearound{'s approach to HOAS using catamorphisms and
  anamorphisms}\citet{Washburn-Weirich-boxes}.

>>>>

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
In contrast to \S\ref{intro}, in the rest of the paper 
we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensures that our object programs are closed.
We find HOAS to be more convenient, but we have also implemented our
approach using de Bruijn indices
(in Section~\ref{S:de-Bruijn} and the accompanying source file
|incope-dB.ml|).

@@

\subsection{Higher-order abstract syntax versus de Bruijn indices}
\label{S:de-Bruijn}

\begin{figure}
\begin{floatrule}
\begin{BVerbatim}
module type Symantics = sig
  type ('c,'h,'dv) repr
  type ('c,'dv) vr			(* variable representation *)

  val vz  : ('c, ('c,'d) vr * 'h, 'd) repr
  val vs  : ('c, 'h, 'd) repr -> ('c, _ * 'h, 'd) repr

  val int : int  -> ('c,'h,int) repr
  val bool: bool -> ('c,'h,bool) repr

  val lam : ('c, ('c,'da) vr * 'h, 'db) repr -> ('c,'h,'da->'db) repr
  val app : ('c,'h,'da->'db) repr -> ('c,'h,'da) repr -> ('c,'h,'db) repr
  val fix : ('c, ('c,'da->'db) vr * 'h, 'da->'db) repr
            -> ('c, 'h, 'da->'db) repr

  val add : ('c,'h,int) repr -> ('c,'h,int) repr -> ('c,'h,int) repr
  val mul : ('c,'h,int) repr -> ('c,'h,int) repr -> ('c,'h,int) repr
  val leq : ('c,'h,int) repr -> ('c,'h,int) repr -> ('c,'h,bool) repr
  val if_ : ('c,'h,bool) repr
            -> (unit -> 'x) -> (unit -> 'x) -> (('c,'h,'da) repr as 'x)
end

module R = struct
  type ('c,'h,'dv) repr = 'h -> 'dv
  type ('c,'d) vr = 'd

  let vz   (x,_) = x
  let vs v (_,h) = v h

  let int  (x:int)  h = x
  let bool (b:bool) h = b
  let lam  f        h = fun x -> f (x,h)
  let app  e1 e2    h = (e1 h) (e2 h)
  let fix  f        h = let rec self n = f (self,h) n in self
  let add  e1 e2    h = e1 h + e2 h
  let mul  e1 e2    h = e1 h * e2 h
  let leq  e1 e2    h = e1 h <= e2 h
  let if_  eb et ee h = if eb h then et () h else ee () h
end
\end{BVerbatim}
\end{floatrule}
\caption{Embedding and evaluating our object language using de Bruijn indices}
\label{fig:de-Bruijn}
\end{figure}

Because Haskell and ML allow case analysis on $\lambda$-bound variables,
one might worry that our HOAS representation of the object language
allows \emph{exotic terms} and is thus inadequate.  To the contrary,
because the representation of an object term is \emph{parametrically}
polymorphic over the type constructor |repr| of the interpreter,
$\lambda$-bound object variables cannot be case-analyzed.
We thus follow \citet{washburn-boxes-jfp} in ``enforcing term
parametricity with type parametricity'' to represent and fold over
abstract syntax.

Although the rest of this paper continues to represent binding
using HOAS\@, our approach is compatible with de Bruijn indices.
The accompanying source file |incope-dB.ml| implements this
alternative, starting with the |Symantics| signature and the |R|
evaluator in Fig.~\ref{fig:de-Bruijn}.

@@

Our encoding of the type function |repr_pe| in \S\ref{S:PE-solution}
emulates type-indexed types and is related to intensional type analysis
\citep{Morrisett-intensional,Generic-Haskell}. However, our object
language and running examples in HOAS include |fix|,
which intensional type analysis cannot handle
\citep{xi-guarded}.
